{\bf \Large TO BE RENAMED}

### 2023-05-31 draft.tex
applying Zak's changes 1

    This thesis comprises of four studies investigating the context effects arising from multidimensional, multi-attribute choice settings and from consumer preferences and applies the knowledge to enhance recommender systems in online marketplaces. My findings indicate that context effects in choice settings which were primarily studied and observed in experimental settings before are also present and detectable in field settings. Moreover, tools that enable users to signal their preferences to recommender systems can significantly increase their adoption. The empirical tests of these essays rely on four distinct and unique datasets, the biggest being an observational data from a field setting and the remaining three coming from an experimental setting.

    The first essay applies a computational decision making model, MDbS, to a substantial real-world choice dataset, making the first application of this magnitude. Findings show that the context effects which have been commonly observed and accounted for in controlled experiments can also apply in field settings. The study suggest that online marketplaces could utilize such models to gain further insights into how the composition of the choice set and interplay among different options affect consumer decisions. This study also has certain limitations related to the capacity of computational models to identify context effects only in aggregate. The capability to recognize these effects in isolation can foster a more profound comprehension of the manifestation of context effects, offering more granular insights into how they individually influence the choice behaviors of individuals when co-existing with other effects.

    Building on the findings of the previous work, the second essay aims to develop a methodology which can disentangle the context effects and separate big three effects known as attraction, compromise and similarity from one another. The unique approach of this study allows further applications of this methodology to online and offline environments. The findings of this study lays the foundation of future research on using contextual information extracted from the choice set into a design of recommender systems with a particular application of tackling the so called cold-start problem.

    The third study presented a novel approach to addressing the user-side cold-start problem in recommender system design. It applied the findings of two-stage choice process observed in individuals to contextual information extracted from the choice set via clustering to generate consideration sets. The approach proved to be robust to clustering method and different heuristics applied. Its findings pave the way for investigating context effects arising from outside the choice sets, namely the preferences of individuals and the user control tools that allows individuals to signal their preferences in designing. This is argued to be one of the main mechanisms to create more effective recommender systems.

@@

\subsection{Purpose and research questions}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{staticFiles/attractionEffect.png}
    \caption{Attraction effect in ternary choice. sOME MORE EXPLANATION} % Add your description here
    \label{fig:attractionEffect} % This labels your figure for reference

@@

This section will discuss the methodology I have used for this study. After brief discussion of the dataset I will move onto detailed discussion of the steps taken in the analysis. SOME MORE BLABLABLA

@@

These are the attributes that are designated as vertical in the choice process. For the purposes of this chapter, it is  assumed that consumers prefer lower values for each of them (e.g., all consumers prefer lower prices, shorter trips, fewer layovers, and not having to change airlines too frequently). Besides vertical attributes, the data also contains two sets of horizontal attributes, the departure times and the dates of out- and in-bound flights. Those attributes are treated as horizontal as we have no clear way of defining consumer preferences over them. To eliminate potential scale effects, z-score normalization on vertical attributes was performed following  $Z = \frac{{x - \overline{x}}}{{\sigma}}
$ where $\overline{x}$ is the mean and $\sigma$ is the standard deviation of variable $x$.

@@

At first sight, such model behavior might seem surprising. Recall that the nature of MDbS is in comparing alternatives with each other and collect evidence based on won comparisons. In the choice experiments, usual size of the menu is one and only one of the context effects is generated at given time. In observational data however, the number of alternatives in the menu is much higher. The presence of large number of alternatives potentially introduces other context effects. Also, MDbS is bound to only to dimensions which are mathematically comparable with each other in the sense of bigger better, or smaller better. In the observational dataset there present also horizontal attributes for which only the decision maker can decide in a given scenario whether or not given the same price, flight duration the flight which is at 5:00 in the morning is better from the one that is at 14:00 afternoon.

@@

On the other hand, there exists another set of attributes where there is no obvious, homogeneous ordering. For example, consider color. We have no theoretical ground to assume that all consumers would prefer a car that is blue, over a car that is green (all other attributes held constant). Same is true about attributes which at first sight are not strictly labelled as categorical, for example time. When buying a cinema or plane ticket there is no theoretical reason for one to explain how a ticket for 15:00 is better or worse than the one at 17:00.  I refer to these as horizontal attributes. Potential heterogeneity across decision-makers in ordering over categories in such attributes makes inclusion of such features in the calculation of attraction and compromise effects impossible. In experimental settings, it is often the case that these attributes are constant across treatments to prevent any confounding effects. However, in the field this usually cannot be done. Therefore, the study of context effects with observational data requires controlling for them statistically.

@@

An important drawback of the dataset is the feature that there is no way of knowing which available options on the market reached eyeballs of the consumer. One way of thinking about this problem is to consider the most likely way such menus are delivered to decision-makers. Most online booking sites and flight aggregators use specific and proprietary algorithms for ranking available menus at the point of search. These rankings decide which options are shown to the customer. Even though the information about specific ranking algorithms is not public, we know that the attribute that usually plays the most important role is the price. Given the robust findings that price negatively affects choice probability, the best guess for a simple ranking mechanism that would capture a wide variety of sorting mechanisms would be options sorted with the decreasing order with respect to price. Assuming that each user was reached by the same number of options, we could construct a reduced dataset for a sensitivity check. For this exercise I construct a dataset that only contains menus with more than 20 options, and only retain 20 cheapest alternatives per menu. There are also cases where the chosen option is not part of the set of 20 cheapest alternatives in the menu \footnote{20 alternatives are chosen so that we have enough entries to have variance in key variables. Results are robust to different menu sizes, with the characteristics that as I reduce menu size more effects seem to lose significance.}. I also eliminate these choice cases from the reduced dataset. This leaves me with about four thousand choice cases.

@@

    \begin{tabular}{>{\fontsize{10pt}{11pt}\selectfont}p{3cm}>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l}
        \hline
        Panel A & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 & Model 8 & Model 9 \\ \hline
        Attraction & 0.0138 & & 0.0191 & & 0.0010 & & \\ 
         & [0.512] & & [0.362] & & [0.317] & & \\
        Compromise & & -0.5507 & & -0.5239 & & -0.5335 & \\ 
         & & [$<0.001$] & & [$<0.001$] & & [$<0.001$] & \\ 
        Similarity & & & -0.1916 & -0.1890 & -0.2425 & -0.1859 & -0.2375 \\ 
         & & & [$<0.001$] & [$<0.001$] & [$<0.001$] & [$<0.001$] & [$<0.001$] \\ 
        \renewcommand{\arraystretch}{1.}
        Attraction within cluster & & & & & 0.1357 & & 0.1329 \\ 
         & & & & & [0.001] & & [0.001] \\ 
        Attraction outside cluster & & & & & -0.0111 & & -0.0080 \\ 
         & & & & & [0.628] & & [0.728] \\ 
        \renewcommand{\arraystretch}{1.}
        Compromise within cluster & & & & & & -1.5127 & -1.4961 \\ 
         & & & & & & [$<0.001$] & [$<0.001$] \\ 
        Comrpmise outside cluster & & & & & & -0.2374 & -0.2566 \\ 
         & & & & & & [0.107] & [0.083] \\ \hline
    \end{tabular}

@@

The data comes from a discrete choice experiment administered to residents and daily commuters to the city of Ljubljana, Slovenia by Gerzinic et al. \citeyearonly{gerzinicEtAl21}. 108 subjects were sequentially presented with 12 5-alternative menus and were asked to choose the best alternative in each case \footnote{In fact, the experiment consisted of multiple choices rounds within each menu. In the first round subjects needed to choose the best option of the five presented alternatives. Consequently, this alternative was removed from the menu and in the second round they needed to choose the worst option out of four remaining alternatives. Then this option was removed and subjects needed to choose the best of the three remaining, and ultimately the worst of the two remaining options. The experiment was designed for a different purpose, see Gerzinic et al. \citeyearonly{gerzinicEtAl21}. For the purpose of this study, only the data from the first round of choices is used.}.  This represents 1,296 recorded choice cases. Each alternative described a commuter trip with a 'park and ride facility choice' to the city with respect to following five characteristics: price, car ride duration, public transport ride duration, public transport (average) wait time, and the mode of public transport (either bus or train). Subjects were overwhelmingly Slovenian nationals ($91.67\%$), 58\% female, with mean age of 36 years, ($St. Dev.=12.5$). Further information on education, income, household size and the number of cars in the household was also obtained. Descriptive  statistics of choice variables in the experimental dataset is presented in table \ref{tab:descriptivesNejc}. Subject characteristics are used as control variables.

@@

It is essential to base this study in a theoretical framework. As previously mentioned, we use Technology Acceptance Model by Davis \citeyearonly{davis1985technology}. Currently there are two dominant technology acceptance models used in the literature: Unified Theory of Acceptance and Use of Technology (UTAUT) and Technology Acceptance Model (TAM) \citep{venkatesh2003utaut, davis1985technology}. Both models provide valuable insights in understanding the users' acceptance of information systems. In the following section we will briefly discuss both TAM and UTAUT and argue why we have ultimately selected TAM for our study. Then, hypotheses will be developed.

@@

There exist many clustering methods and each of them have a different way of defining similar items and as a result will group in different ways. Clustering is usually an unsupervised machine learning method in that there are no preconceived labels given to the clusters. This implies that we have no universal way of evaluating the quality of a clustering outcome. Moreover, most clustering methods require an additional input to determine the number of clusters.  Therefore, in order to make sure our results do not depend on the chosen clustering method, I investigate the two widely used methods - affinity propagation and k-means clustering.

>>>>

    This thesis comprises of four studies investigating the context effects arising from multidimensional, multi-attribute choice settings and applies the knowledge to enhance recommender systems in online marketplaces. The findings indicate that context effects in choice settings which were primarily studied and observed in experimental settings before are also present and detectable in field settings. Moreover, tools that enable users to signal their preferences to recommender systems can significantly increase its adoption. The empirical tests of these chapters rely on four distinct and unique datasets, the largest being an observational data from a field setting and the remaining three coming from an experimental setting.

    The first chapter applies a computational decision making model to a substantial real-world choice dataset, making the first application of this magnitude. Findings indicate that context influences individuals' choices. The study suggest that online marketplaces could utilize such models to gain further insights into how the composition of the choice set and interplay among different options affect consumer decisions. The key feature of this computational model is that it identifies context in aggregate. Understanding the underlying components of context can foster a more profound comprehension, offering more granular insights into how each of them individually influence the choice behaviors of individuals when co-existing with other effects.

    Building on the results of the first chapter, the second chapter extends beyond the traditional studies of context, by developing a methodology that aims to disentangle its three main components known as attraction, compromise and similarity from one another. This methodology is designed to be versatile, able to cope with diverse choice settings with varying sizes. Data sourced from both online and offline environments can be successfully integrated into this framework, enhancing its scope and applicability. Insights derived from this study offer potential to fine-tune online product offers and recommender systems. One prominent application of this approach lies in navigating the 'cold-start' problem faced digital marketplaces. It advocates the design of consideration sets that entail empirically similar alternatives, fostering effective decision-making in digitally crowded spaces. This innovative approach represents a stepping stone for future research to extract and utilize contextual information from choice sets. Hence, it contributes to the development of recommender system designs and to the deeper understanding of heterogeneous nature of consumer choice dynamics.

    The third study presented a novel approach to addressing the user-side cold-start problem in recommender system design. It applies the findings of two-stage choice process observed in individuals to contextual information extracted from the choice set via clustering to generate consideration sets. The approach proves to be robust to clustering method and different heuristics applied. Its findings pave the way for investigating context effects arising from outside the choice sets, namely the preferences of individuals and the tools that allow them to signal their preferences to recommender systems. This is argued to be one of the main mechanisms to create more effective recommender systems.

@@

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{staticFiles/attractionEffect.png}
    \caption{Attraction effect in ternary choice.} % Add your description here
    \label{fig:attractionEffect} % This labels your figure for reference

@@

This section will discuss the methodology I have used for this study. After brief discussion of the dataset I will move onto detailed discussion of the steps taken in the analysis.

@@

These are the attributes that are designated as vertical in the choice process. For the purposes of this chapter, it is  assumed that consumers prefer lower values for each of them (e.g., all consumers prefer lower prices, shorter trips, fewer layovers, and not having to change airlines too frequently). Besides vertical attributes, the data also contains two sets of horizontal attributes, the departure times and the dates of out- and in-bound flights. Those attributes are treated as horizontal as there is no clear way of defining consumer preferences over them. To eliminate potential scale effects, z-score normalization on vertical attributes was performed following  $Z = \frac{{x - \overline{x}}}{{\sigma}}
$ where $\overline{x}$ is the mean and $\sigma$ is the standard deviation of variable $x$.

@@

At first sight, such model behavior might seem surprising. Recall that the nature of MDbS is in comparing alternatives with each other and collect evidence based on won comparisons. In the choice experiments, usual size of the menu is three and only one of the context effects is generated at given time. In observational data however, the number of alternatives in the menu is much higher. The presence of large number of alternatives potentially introduces other context effects. Also, MDbS is bound to only to dimensions which are mathematically comparable with each other in the sense of bigger better, or smaller better. In the observational dataset there present also horizontal attributes for which only the decision maker can decide in a given scenario whether or not given the same price, flight duration the flight which is at 5:00 in the morning is better from the one that is at 14:00 afternoon.

@@

On the other hand, there exists another set of attributes where there is no obvious, homogeneous ordering. For example, consider color. There is no theoretical ground to assume that all consumers would prefer a car that is blue, over a car that is green (all other attributes held constant). Same is true about attributes which at first sight are not strictly labelled as categorical, for example time. When buying a cinema or plane ticket there is no theoretical reason for one to explain how a ticket for 15:00 is better or worse than the one at 17:00.  I refer to these as horizontal attributes. Potential heterogeneity across decision-makers in ordering over categories in such attributes makes inclusion of such features in the calculation of attraction and compromise effects impossible. In experimental settings, it is often the case that these attributes are constant across treatments to prevent any confounding effects. However, in the field this usually cannot be done. Therefore, the study of context effects with observational data requires controlling for them statistically.

@@

An important drawback of the dataset is the feature that there is no way of knowing which available options on the market reached eyeballs of the consumer. One way of thinking about this problem is to consider the most likely way such menus are delivered to decision-makers. Most online booking sites and flight aggregators use specific and proprietary algorithms for ranking available menus at the point of search. These rankings decide which options are shown to the customer. Even though the information about specific ranking algorithms is not public, we know that the attribute that usually plays the most important role is the price. Given the robust findings that price negatively affects choice probability, the best guess for a simple ranking mechanism that would capture a wide variety of sorting mechanisms would be options sorted with the decreasing order with respect to price. Assuming that each user was reached by the same number of options, we could construct a reduced dataset for a sensitivity check. For this exercise I construct a dataset that only contains menus with more than 20 options, and only retain 20 cheapest alternatives per menu. There are also cases where the chosen option is not part of the set of 20 cheapest alternatives in the menu \footnote{20 alternatives are chosen so that there are enough entries to have variance in key variables. Results are robust to different menu sizes, with the characteristics that as I reduce menu size more effects seem to lose significance.}. I also eliminate these choice cases from the reduced dataset. This leaves me with about four thousand choice cases.

@@

    \begin{tabular}{>{\fontsize{10pt}{11pt}\selectfont}p{3cm}>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l>{\fontsize{11pt}{13pt}\selectfont}l}
        \hline
        Panel A & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 & Model 8 & Model 9 \\ \hline
        Attraction & 0.0138 & & 0.0191 & & 0.0010 & & \\ 
         & [0.512] & & [0.362] & & [0.317] & & \\
        Compromise & & -0.5507 & & -0.5239 & & -0.5335 & \\ 
         & & [$<0.001$] & & [$<0.001$] & & [$<0.001$] & \\ 
        Similarity & & & -0.1916 & -0.1890 & -0.2425 & -0.1859 & -0.2375 \\ 
         & & & [$<0.001$] & [$<0.001$] & [$<0.001$] & [$<0.001$] & [$<0.001$] \\ 
        \renewcommand{\arraystretch}{1.}
        Attraction within cluster & & & & & 0.1357 & & 0.1329 \\ 
         & & & & & [0.001] & & [0.001] \\ 
        Attraction outside cluster & & & & & -0.0111 & & -0.0080 \\ 
         & & & & & [0.628] & & [0.728] \\ 
        \renewcommand{\arraystretch}{1.}
        Compromise within cluster & & & & & & -1.5127 & -1.4961 \\ 
         & & & & & & [$<0.001$] & [$<0.001$] \\ 
        Comrpmise outside cluster & & & & & & -0.2374 & -0.2566 \\ 
         & & & & & & [0.107] & [0.083] \\
    \end{tabular}

@@

The data comes from a discrete choice experiment administered to residents and daily commuters to the city of Ljubljana, Slovenia by Gerzinic et al. \citeyearonly{gerzinicEtAl21}. 108 subjects were sequentially presented with 12 5-alternative menus and were asked to choose the best alternative in each case \footnote{In fact, the experiment consisted of multiple choice rounds within each menu. In the first round subjects needed to choose the best option of the five presented alternatives. Consequently, this alternative was removed from the menu and in the second round they needed to choose the worst option out of four remaining alternatives. Then this option was removed and subjects needed to choose the best of the three remaining, and ultimately the worst of the two remaining options. The experiment was designed for a different purpose, see Gerzinic et al. \citeyearonly{gerzinicEtAl21}. For the purpose of this study, only the data from the first round of choices is used.}.  This represents 1,296 recorded choice cases. Each alternative described a commuter trip with a 'park and ride facility choice' to the city with respect to following five characteristics: price, car ride duration, public transport ride duration, public transport (average) wait time, and the mode of public transport (either bus or train). Subjects were overwhelmingly Slovenian nationals ($91.67\%$), 58\% female, with mean age of 36 years, ($St. Dev.=12.5$). Further information on education, income, household size and the number of cars in the household was also obtained. Descriptive  statistics of choice variables in the experimental dataset is presented in table \ref{tab:descriptivesNejc}. Subject characteristics are used as control variables.

@@

It is essential to base this study in a theoretical framework. As previously mentioned, we use Technology Acceptance Model by Davis \citeyearonly{davis1985technology}. Currently there are two dominant technology acceptance models used in the literature: Unified Theory of Acceptance and Use of Technology (UTAUT) and Technology Acceptance Model (TAM) \citep{venkatesh2003utaut, davis1985technology}. Both models provide valuable insights in understanding the users' acceptance of information systems. In the following section we will briefly discuss both TAM and UTAUT and argue why I have ultimately selected TAM for our study. Then, hypotheses will be developed.

@@

There exist many clustering methods and each of them have a different way of defining similar items and as a result will group in different ways. Clustering is usually an unsupervised machine learning method in that there are no preconceived labels given to the clusters. This implies that there is no universal way of evaluating the quality of a clustering outcome. Moreover, most clustering methods require an additional input to determine the number of clusters.  Therefore, in order to make sure our results do not depend on the chosen clustering method, I investigate the two widely used methods - affinity propagation and k-means clustering.
